Citation Analysis of Computer Science Publications

### 2019-12-04 Scientometrics/comp.tex
update

\begin{abstract}
Computer science as a field has experienced dramatic growth and diversification over the last twenty years. Towards an improved understanding of the structure of the field, we analyze a cohort of the computer science literature.  For insight on the features of this cohort and the flow of information within its components, we construct article level clusters whose elements are linked by citations or co-citations, and reconcile them with subject categories in the All Science Journal Classification (Scopus). [abstract is feeble and needs to be rewritten]

@@

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. The collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have further been postulated as a recent driver for its growth and evolution~\cite{siebel2019_digital}. As also noted in a 2017 US National Academies Report~\cite{nas_2017}, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission". Given this powerful influence of computer science in research, some understanding of its present state, structure, and relationship to other fields that seems warranted since it can inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of its broad influence, computer science can also be an excellent model system for studying interdisciplinarity. These are not novel ideas, however, as far back as 1979, Salton and Bergmark conducted a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles) in which they made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas described were  (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus. the Web of Science research and categorical classifications, the NSF classification system~\cite{nsf_classification}. Of these three, Scopus and the Web of Science make also contain citation data and proprietary unique identifiers for publications that can be matched, in some cases, to digital object identifiers (DOIs). All three rely on applying one or more journal-derived labels to articles. Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been explored~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}, which escape some of the criticisms of journal-based classification.

@@

Thus, we sought to extend the work of Salton and Bergmark to reflect a more current reality while also considering connections between computer science and other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage is incomplete and possibly biased. Cross-matching DBLP publications to Scopus records using whole text searches applied to titles, enabled harvesting citation data as well as publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus as well as to the CCS classification through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

@@

\paragraph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments.
The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus. We varied the number of clusters to be formed between 10 and 50 clusters. At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}. Consequently, we generated 20 clusters using Graclus, which were labeled 0-19 (Fig~\ref{fig:heatmap})

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we used variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size returned to the process and a higher threshold applied to break such clusters.

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 634,100 articles in \emph{comp}. After discarding those publications without any citations, we restricted further analysis to those in the 90th percentile of citations or higher resulting in a dataset of size 50,869. We then identified 897,407 publications in Scopus that cite the 50,869 previously selected papers from \emph{comp}. The cited references in these papers were restricted to those in the set of 50,869 highly cited papers previously identified. For each of the 897,407 citing publications in turn, all possible reference pairs, ${n \choose 2}$, were generated where $n$ is the number of references in a publication. The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 2,208,135 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = 0.5 (normalized co-citation frequency, see Methods), increment $i = 0.1$, and maximum cluster size, $mcs = 100$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 6,822 clusters containing 35,485 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 3,851. Pairwise clustering was then performed on these clusters to generate higher-order clusters. After 9 rounds of pairwise clustering, 13 clusters were generated resulting from agglomeration of 3,851 clusters

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus3.pdf}
% figure caption is below the figure
\caption{caption will be inserted I just haven't been able to overcome the threshold to avoid boring tasks}
\label{fig:heatmap}       % Give a unique label
\end{figure}
For the purpose of this study, our working definition of the computer science literature was all publications in the DBLP bibliography published in the years 1996-2015, that had a digital object identifier (DOI), and could also be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 634,100 publications (Materials and Methods). 
INSERT sentence about why one should cluster this dataset. Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using a spectral clustering technique, Graclus, that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. We constructed article level clusters of this computer science dataset at a sufficiently high level of aggregation to avoid cognitive challenge and cross-matched them to disciplinary classification by journal, the Scopus ASJC classification. We identified ASJC categories that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 minor subject areas in the Scopus ASJC classification, which consists of 5 subject areas, 27 major subject areas, and 334 minor subject areas (Fig~\ref{fig:heatmap}).

Unsurprisingly, the category Computer Science Applications was present in 17 of 20 clusters, followed by  Electrical and Electronic Engineering in 12 clusters. At the other extreme, 16 minor subject areas were limited to a single cluster, viz.,
Atomic and Molecular Physics and Optics; 
Biochemistry; 
Chemistry (all); 
Cognitive Neuroscience; 
Computer Vision and Pattern Recognition;
Computers in Earth Sciences;
Earth and Planetary Sciences; 
Education; 
Electronic Optical and Magnetic Materials; 
Information Systems; 
Information Systems and Management;  
Management Science and Operations Research;
Molecular Biology;
Neurology; 
Physics and Astronomy (miscellaneous); 
and Radiology Nuclear Medicine and Imaging. These 16 areas can be described as peripheral, in various degrees, to computer science. The picture that emerges is that areas central to computer science are likely to be found in multiple clusters. We did not attempt to map these clusters to Salton and Bergmark's classification.

Examining these data from the perspective of the generated clusters (Fig. 1 abscissa), interactions with the Life Sciences were evident in cluster 17 (Biochemistry, Computational Mathematics, Computational Theory and Mathematics, Computer Science Applications, and Molecular Biology) and cluster 10 (Artificial Intelligence, Biomedical Engineering, Cognitive Neuroscience, Computer Science Applications, and Neurology). While acknowledging previously reported inaccuracies in journal-based classification \emph{vide supra}, these results strongly suggest that the actual structure of communication within the broad area of computer science is at variance with journal-based classification, also noted by Waltman and van Eck, to wit, ``...an indication that traditional disciplines such as those just mentioned only partly reflect the actual organization of today's scientific research."~\cite{waltman_new_2012}. The aphorism of Porter, Cohen, Roessner, and Perreault (2007), `The discrepancies between disciplinary roots and research realities loom large'~\cite{porter_measuring_2007} also invokes itself. Secondly, that a correlation exists between broad journal-based classification and article clusters in terms of subject matter. Thirdly, that some areas of computer science are relatively ubiquitous. Fourth that interactions with fields outside computer science tend to be restricted to a single cluster or small numbers of clusters. We also attempted to map these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that this small sample (top cited papers) were not representative of the cluster.

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.3]{vlc.pdf}
% figure caption is below the figure
\caption{Schematic representation of variable clustering protocol modified from Small and Sweeney (1985). Three parameters are specified (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. Input data is a set of co-cited publications with edge-weight defined by normalized co-citation frequencies. Green clusters are within the max cluster size. At the initial threshold, $t1$, a single cluster below the maximum cluster size, $mcs$ (green), along with one large cluster above it (red) are generated. As the threshold is incremented to $t2$, additional clusters of acceptable size is generated. The cascade continues to completion, which is defined by all clusters being of size less than or equal to the $mcs$. In this schematic, five rounds are adequate for the process to run to completion.}
\label{vlc_process}       % Give a unique label
\end{figure}

>>>>

\begin{abstract}
Computer science as a field has experienced dramatic growth and diversification over the last twenty years. Towards an improved understanding of the structure of the field, we analyze a cohort of the computer science literature.  For insight on the features of this cohort and the flow of information within its components, we construct article level clusters whose elements are linked by citations or co-citations, and reconcile them with major and minor subject categories in the All Science Journal Classification (Scopus). [abstract is feeble and needs to be rewritten]

@@

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. More recently, the collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have been postulated as a recent driver for growth and evolution~\cite{siebel2019_digital}. As noted in a 2017 US National Academies Report~\cite{nas_2017}, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission". Given the powerful influence of computer science in research, some understanding of its present state, structure, and relationship to other fields that seems warranted since it can inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of broad influence, computer science can also be an excellent model system for studying interdisciplinarity.

These are not novel ideas however, as far back as 1979, Salton and Bergmark conducted a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles) in which they made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas noted were  (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus. the Web of Science research and categorical classifications, the NSF classification system~\cite{nsf_classification}. Of these, Scopus and the Web of Science also contain citation links and proprietary unique identifiers for publications that can be matched, in some cases, to digital object identifiers (DOIs). All three rely on applying one or more journal-derived labels to articles. However, Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been explored~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}, which escape some of the criticisms of journal-based classification.

@@

We sought to extend the work of Salton and Bergmark to reflect a more current reality while also considering connections between computer science and other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage is incomplete and possibly biased. Cross-matching DBLP publications to Scopus records using DOIs, enabled of harvesting citation data as well as links to publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

@@

\paragraph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments.
The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus. We varied the number of clusters to be formed between 10 and 50 clusters. At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}. Consequently, we generated 20 clusters, labeled 0-19, using Graclus (Fig~\ref{fig:heatmap}). The quality of clustering was evaluated by measuring conductance [add citations] as defined by Shun.

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we first implemented variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size returned to the process and a higher threshold applied to break such clusters.

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, we restricted further analysis to those in the 90th percentile of citations or higher resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite the 212,311 previously selected papers from \emph{comp}. The cited references in these papers were restricted to those in the set of 212,311 highly cited papers previously identified. For each of the 4,318,305 citing publications in turn, all possible reference pairs, ${n \choose 2}$, were generated where $n$ is the number of references in a publication. The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 46,463,117 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs = 200$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 22,232 clusters containing 84,591 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these clusters to generate higher-order clusters. Briefly, the edge weight between two clusters was assigned as the maximum edge weight of all edges between these two clusters. Edges were arranged in descending order and the first pair of clusters were merged and its edge weight with other interacting clusters was recalculated. The edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (figure), and resulted in 665 clusters. The top 20 clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus3.pdf}
% figure caption is below the figure
\caption{caption will be inserted I just haven't been able to overcome the threshold to avoid boring tasks}
\label{fig:heatmap}       % Give a unique label
\end{figure}
For the purpose of this study, our working definition of the computer science literature was all publications in the DBLP bibliography published in the years 1996-2015, that had (i) a digital object identifier (DOI), and (ii) could also be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 2,685,356 publications (Materials and Methods).

[Need justification for why 20 article level clusters- may be in Waltman's article]

Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using Graclus, a spectral clustering technique that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. We constructed article level clusters of this computer science dataset at a sufficiently high level of aggregation to avoid cognitive challenge and cross-matched them to disciplinary classification by journal, the Scopus ASJC classification. We considered Scopus ASJC minor subject area categories that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 ASJC minor subject areas, suggesting some selectivity in these results.

(Fig~\ref{fig:heatmap}) permits examination of these data from two perspectives (i) the clusters that map to a given ASJC minor subject area (ii) ASJC minor subject areas that comprised at least 10\% of the publication in a cluster. Unsurprisingly, the broad category Computer Science Applications was present in 16 of 20 clusters, followed by Software in 12 clusters, and Electrical and Electronic Engineering in 10 clusters.

From the other perspective, cluster 17 was the most diverse and contained publications annotated with 8 minor subject area labels, of which Biochemistry, Chemistry(all), Genetics, and Molecular Biology were the most distant from computer science while Statistics \& Probability, Computational Theory \& Mathematics, and Computational Mathematics are closer. It is important to be reminded of the caveats associated with (i) the limitations of the DBLP dataset (ii) the limitations of cross-matching to Scopus (iii) the limitations of journal based classification (iv) the limitations of clustering by direct citation. Even so, a picture that emerges is that areas central to computer science in 2019 are more likely to be found in multiple clusters (hardware, software, and theory) and that biology, chemistry, and statistics are proximal.  These results strongly suggest that the actual structure of communication within the broad field of computer science may be at variance with journal-based classification, also noted by Waltman and van Eck, ``...an indication that traditional disciplines such as those just mentioned only partly reflect the actual organization of today's scientific research."~\cite{waltman_new_2012}. The aphorism of Porter, Cohen, Roessner, and Perreault (2007), `The discrepancies between disciplinary roots and research realities loom large'~\cite{porter_measuring_2007} also invokes itself. A second inference is that a correlation appears between broad journal-based classification and article clusters in terms of subject matter. Thirdly, that some areas of computer science are relatively ubiquitous. Fourth that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters.

We also attempted to map these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that this small sample (top cited papers) were not representative of the cluster.

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.25]{vlc.pdf}
% figure caption is below the figure
\caption{Schematic representation of variable clustering protocol modified from Small and Sweeney (1985). Three parameters are specified (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. Input data is a set of co-cited publications with edge-weight defined by normalized co-citation frequencies. Green clusters are within the max cluster size. At the initial threshold, $t1$, a single cluster below the maximum cluster size, $mcs$ (green), along with one large cluster above it (red) are generated. As the threshold is incremented to $t2$, additional clusters of acceptable size is generated. The cascade continues to completion, which is defined by all clusters being of size less than or equal to the $mcs$. In this schematic, five rounds are adequate for the process to run to completion.}
\label{vlc_process}       % Give a unique label
\end{figure}
